---
title: "Salt Chemotaxis"
author: "Vivian Vy Le"
date: '2022-03-17'
output: 
  prettydoc::html_pretty:
    toc: TRUE
    theme: hpstr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Guanylyl cyclases (*gcy*) are enzymes that can convert GTP to cGMP. These proteins are necessary for regulating the cGMP-gated cation channels. The regulation of cGMP-gated cation channels helps produce calcium currents, which helps nematodes chemotaxis and detect salt concentrations in the environment.

While all the information on chemosensation is in the *Caenorhabditis elegans*, there is an interest in studying salt chemosensation and the gcy gene family in host associated and parasitic nematodes. *Pristionochus pacifcius* is a non-parasitic nematode species that has a host assocation towards the Oriental beetle, and is attracted to a beetle sex pheromone.

It is unclear if the *gcy* genes in *P.pacificus* have similar gene functions in chemotaxis towards water soluble salts and are expressed in the amphid sensory neuron. With *P. pacificus*' host association towards the Oriental beetle, it would be interesting to study if cGMP signaling contributes to the species' host-seeking abilities.

## Load libraries

```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(palettetown)
library(LaCroixColoR)
library(magick)
```

### Load data

```{r}
salt <- read.csv(here("Data", "salt_CTX_data.csv"))
allgenes <- read.csv(here("Data", "salt_ctx_all_genes.csv"))
```

### View data

```{r}
glimpse(salt)
glimpse(allgenes)
```

# Method

Chemosensation can be quantified by attraction and repulsion. To measure responses, we established a salt gradient on media.

<img src = "https://raw.githubusercontent.com/vivvle/Salt_Chemotaxis/main/Photos/template_labeled.png">


A represents attractant. Salts are considered attractants for the purpose of the experiment.

The list of salts used are the following:

-   NH4Br
-   NH4Cl
-   NH4I
-   LiCl
-   NaCl
-   NaAc

C represents the counter-attractant. We used water as the counter-attractant.

We wash and place the worms on the loading zone of the plate. Once the worms are placed onto the plate, the assay starts and the worms move for 3-5 hours before counting.

# Data analysis

The raw data consists of the number of worms in the attractant zone, the number of worms in the counter-attractant zone, and the number of worms dispersed from the loading zone. I calculated the chemotaxis index (CI) for the data set.


## gcy-22 mutants 
```{r}
salt_data <- salt %>%
  mutate(total = (A+C)) %>% #adding A and C for total worm participating in the assay
  mutate(CI = ((A-C)/(total))) %>% #calculating for the chemotaxis index with the worms participating in the assay
  mutate(participation = (total)/(total+Dispersed)) %>% #calculating for the fraction of worm participating from the population
  relocate("total", .before = "Dispersed") %>% #moved total column before dispersed
  mutate(Strain = factor(Strain,
                         levels = c("wildtype (PS312)", "gcy-22 (csu79)", "gcy-22 (csu80)"))) %>%
  mutate(Salt = factor(Salt,
                       levels = c("NH4Br", "NH4Cl", "NH4I", "LiCl", "NaCl", "NaAc")))

head(salt_data)


gcy_data <- salt_data %>%
  select(Strain, Salt, CI) %>%
  group_by(Strain, Salt) %>%
  summarise(average = mean(CI, na.rm = TRUE), #calculating for average
            st_dev = sd(CI, na.rm = TRUE)) #calculating for standard deviation
gcy_data
```
### Plot for gcy-22 mutants
```{r}
plotbase <- function (data, x, y){
  ggplot(data = data, aes(x = Salt, y = average, fill = Strain)) + #creating a plot function for the data
    geom_col(position = "dodge") +
    ylim(-1,1) + #setting y axis limits for graph
    labs(title = "Salt Chemotaxis",
         x = "Salts",
         y = "Chemotaxis Index (CI)")
}

plotbase(gcy_data, x, y) +
  scale_fill_manual(values = pokepal(3)) +
  labs(subtitle = "gcy-22 mutants") #subtitle indicates which strains are on the plot
```


## All mutants
```{r}
allgenes_data <- allgenes %>%
  mutate(total = (A+C)) %>% #adding for total worms pariticpating in the assay
  mutate(CI = (A-C)/(total)) %>% #calculating for CI values
  mutate(participation = (total)/(total+Dispersed)) %>%
  relocate("total", .before = "Dispersed") %>%
  mutate(Salt = factor(Salt,
                       levels = c("NH4Br", "NH4Cl", "NH4I", "LiCl", "NaCl", "NaAc"))) %>%
  mutate(Strain = factor(Strain,
                         levels = c("wildtype (PS312)", "gcy-22 (csu79)", "gcy-22 (csu80)", 
                                    "che-1 (ot5012)", "che-1 (ot5013)", "daf-11 (mok83)", 
                                    "daf-11 (mok84)", "tax-2 (mok59)")))

head(allgenes_data)


genes_ctx <- allgenes_data %>%
  group_by(Strain, Salt) %>%
  summarise(average = mean(CI, na.rm = TRUE), #calculating for averages
            st_dev = sd(CI, na.rm = TRUE)) #calculating for standard deviation

genes_ctx

```

### Separating by gene and strains
```{r}
daf<-genes_ctx %>% #separating for daf-11
  filter(Strain == "wildtype (PS312)" | Strain == "daf-11 (mok83)" | Strain == "daf-11 (mok84)")
daf

che1 <-genes_ctx %>% #separting for che-1
  filter(Strain == "wildtype (PS312)" | Strain == "che-1 (ot5012)" | Strain == "che-1 (ot5013)")
che1

tax <-genes_ctx %>% #separating for tax-2
   filter(Strain == "wildtype (PS312)" | Strain == "tax-2 (mok59)")
tax
```
## Plot for daf-11
```{r}
plotbase(daf, x, y) +
  scale_fill_manual(values = pokepal(3)) +
  labs(subtitle = "daf-11 mutants")
```
## Plot for che-1
```{r}
plotbase(che1, x, y) +
  scale_fill_manual(values = pokepal(3)) +
  labs(subtitle = "che-1 mutants")
```
## Plot for tax-2
```{r}
plotbase(tax, x, y) +
  scale_fill_manual(values = pokepal(3)) +
  labs(subtitle = "tax-2 mutant")
```


## Plot for all mutants
```{r}
plotbase(genes_ctx, x, y) +
  scale_fill_manual(values = pokepal(3)) +
  labs(subtitle = "Overview of genes involved in chemosensation")
```