---
title: "Salt Chemotaxis"
author: "Vivian Vy Le"
date: '2022-03-17'
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Guanylyl cyclases (*gcy*) are enzymes that can convert GTP to cGMP. These proteins are necessary for regulating the cGMP-gated cation channels. The regulation of cGMP-gated cation channels helps produce calcium currents, which helps nematodes chemotaxis and detect salt concentrations in the environment.

While all the information on chemosensation is in the *Caenorhabditis elegans*, there is an interest in studying salt chemosensation and the gcy gene family in host associated and parasitic nematodes. *Pristionochus pacifcius* is a non-parasitic nematode species that has a host assocation towards the Oriental beetle, and is attracted to a beetle sex pheromone.

In *C. elegans*, there are a total of 34 *gcy* genes and are expressed in the amphid sensory neuron ASE. It is unclear if the *gcy* genes in *P.pacificus* have similar gene functions in chemotaxis towards water soluble salts and are expressed in the amphid sensory neuron. With *P. pacificus*' host association towards the Oriental beetle, it would be interesting to study if cGMP signaling contributes to the species' host-seeking abilities.

## Load libraries

```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(palettetown)
library(LaCroixColoR)
library(magick)
```

### Load data

```{r}
salt <- read.csv(here("Data", "salt_CTX_data.csv"))
allgenes <- read.csv(here("Data", "salt_ctx_all_genes.csv"))
```

### view data

```{r}
glimpse(salt)
glimpse(allgenes)
view(salt)
view(allgenes)
```

## Method

Chemosensation can be quantified by attraction and repulsion. To measure responses, we established a salt gradient on media.

<img src = "https://raw.githubusercontent.com/vivvle/Salt_Chemotaxis/main/Photos/template_labeled.png">


A represents attractant. Salts are considered attractants for the purpose of the experiment.

The list of salts used are the following:

-   NH4Br
-   NH4Cl
-   NH4I
-   LiCl
-   NaCl
-   NaAc

C represents the counter-attractant. We used water as the counter-attractant.

We wash and place the worms on the loading zone of the plate. Once the worms are placed onto the plate, the assay starts and the worms move for 3-5 hours before counting.

## Data analysis

The raw data consists of the number of worms in the attractant zone, the number of worms in the counter-attractant zone, and the number of worms dispersed from the loading zone. I calculated the chemotaxis index (CI) for the data set.


### gcy-22 mutants 
```{r}
salt_data <- salt %>%
  mutate(total = (A+C)) %>% #adding A and C for total worm participating in the assay
  mutate(CI = ((A-C)/(total))) %>% #calculating for the chemotaxis index with the worms participating in the assay
  mutate(participation = (total)/(total+Dispersed)) %>% #calculating for the fraction of worm participating from the population
  relocate("total", .before = "Dispersed") %>% #moved total column before dispersed
  mutate(Strain = factor(Strain,
                         levels = c("wildtype (PS312)", "gcy-22 (csu79)", "gcy-22 (csu80)"))) %>%
  mutate(Salt = factor(Salt,
                       levels = c("NH4Br", "NH4Cl", "NH4I", "LiCl", "NaCl", "NaAc")))
salt_data


gcy_data <- salt_data %>%
  select(Strain, Salt, CI) %>%
  group_by(Strain, Salt) %>%
  summarise(average = mean(CI, na.rm = TRUE),
            st_dev = sd(CI, na.rm = TRUE))
gcy_data
```
```{r}
gcy_data %>%
  ggplot(aes(x = Salt, y = average, fill = Strain)) +
  geom_col(position = "dodge") +
  ylim(-1,1) +
  scale_fill_manual(values = pokepal(2)) +
  labs(title = "Salt Chemotaxis for gcy-22",
       x = "Salt",
       y = "Chemotaxis Index (CI)")
```


### All mutants
```{r}
allgenes_data <- allgenes %>%
  mutate(total = (A+C)) %>%
  mutate(CI = (A-C)/(total)) %>%
  mutate(participation = (total)/(total+Dispersed)) %>%
  relocate("total", .before = "Dispersed") %>%
  mutate(Salt = factor(Salt,
                       levels = c("NH4Br", "NH4Cl", "NH4I", "LiCl", "NaCl", "NaAc"))) %>%
  mutate(Strain = factor(Strain,
                         levels = c("wildtype (PS312)", "gcy-22 (csu79)", "gcy-22 (csu80)", 
                                    "che-1 (ot5012)", "che-1 (ot5013)", "daf-11 (mok83)", 
                                    "daf-11 (mok84)", "tax-2 (mok59)")))

allgenes_data

genes_ctx <- allgenes_data %>%
  group_by(Strain, Salt) %>%
  summarise(average = mean(CI, na.rm = TRUE),
            st_dev = sd(CI, na.rm = TRUE))
genes_ctx
```
### plot for all mutants
```{r}
genes_ctx %>% 
  ggplot(aes(x = Salt, y = average, fill = Strain)) +
  geom_col(position = "dodge") +
  ylim(-1,1) +
  scale_fill_manual(values = pokepal(3)) +
  labs(title = "Salt Chemotaxis",
       x = "Salt",
       y = "Chemotaxis Index (CI)")
```

